---
title: "Heat, flood and tornados kill and maim the most people in the US, while flood and hurricanes cause the most damage"
author: "Phillip Rowe"
date: "8/23/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

1. To help local and national US government officials respond to health and economic risks from weather events, we analyzed the data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database of events from 1951-2011, focusing on three dimensions: time, geographic location, and weather event type.
2. For time, the majority of data was collected between 1992 and 2011; thus, we broke our analysis into 5 periods of 4 years each, to smooth for weather events which had large impact but did not occur annually.  
3. We believe a 4-year period is also probably more suitable to the budgeting and planning schedules of government officials than a longer period (such as a decade).
4. Geographic analysis was limited to four time zones: PST, MST, CST, EST. 
5. Weather events were cleaned up (i.e., recategorized) and then filtered to include only twelve types of events, each of which accounted for more than 2% of at least one of the impacts (fatalities, injuries, property damage, or crop damage). 
6. From this analysis, we found that 5 types of weather (heat, flood, wind,  tornadoes, and lightning) accounted for more than 75% of fatalities and injuries in the 20 years analyzed, while flood and hurricanes were the main contributors to property damage. 
7. Flood and drought were the main causes of crop damage.  
8. In addition, a vast majority of fatalities and crop damage and injuries  were concentrated in the Eastern Standard and Central Standard time zones, while  the most property damage was suffered in the Central Standard and Pacific Standard time zones.
9. The Mountain Standard time zone, which was the least impacted across all risks, likely due to lower population density, had a large portion of property damage due to hail, unlike other regions. 
  
## Data Processing 

#### Contents

1. **Load the data and pre-process data:** 
2. **Analyze and limit data by cutoff year and timezone:**
3. **Clean/recategorize weather event types:**
4. **Melt and recast data:**
5. **Time period analysis:** 
6. **Geographic (time zone) analysis:**

##
1. **Load the data and pre-process data:** filter the variables to be those relevant for our analysis, including formatting the bgn_date variable as a date, cleaning damage coding, and adding new variables for **year** and damage in millions of dollars (**propertydamage_M** and **cropdamage_M**).   

``` {r load, cache=TRUE, echo=TRUE, message=FALSE,comment=FALSE}
library("dplyr")
library(lubridate)
library(ggplot2)
library(downloader)
library('stringr')
library('tibble')
library('reshape2')

#---------------------------------------------------------------------------------------
#
# DATA PROCESSING
# 1. **Load the data and pre-process data:** 
#
#---------------------------------------------------------------------------------------

# uncomment following two lines for initial download
url<-'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download(url,dest="storm_data.csv.bz2")  

sdata<-read.csv('storm_data.csv.bz2')

# Fields relevant for health: FATALITIES,INJURIES,
# Fields relevant for economic impact: PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP
# Fields relating to geography: STATE__,TIME_ZONE,STATE
# Other fields: EVTYPE,BGN_DATE
cols<-c('EVTYPE','BGN_DATE','FATALITIES','INJURIES','PROPDMG','PROPDMGEXP',
        'CROPDMG','CROPDMGEXP','TIME_ZONE','STATE__','STATE')

mydf<-select(sdata,cols)
names(mydf)<-tolower(names(mydf))
mydf$evtype<-tolower(mydf$evtype)
mydf$bgn_date<-gsub("\\s0:00:00?","",mydf$bgn_date)
mydf$bgn_date<-as.Date(mydf$bgn_date,"%m/%d/%Y")

# remove rows where the coding for property damage is unclear; there are only 14 of these rows 
# and they do not include any fatalities or injuries, so the impact on our analysis 
# is likely low
prep_propdmgexp<-function(mydf) {
     mydf<-filter(mydf,!(propdmgexp%in%c("?","-","+")))
     mydf<-filter(mydf,!(cropdmgexp%in%c("?")))
     # change coding to be power of ten
     mydf$propdmgexp<-tolower(mydf$propdmgexp)
     mydf$propdmgexp[grep("k",mydf$propdmgexp)]<-3
     mydf$propdmgexp[grep("h",mydf$propdmgexp)]<-2
     mydf$propdmgexp[grep("b",mydf$propdmgexp)]<-9
     mydf$propdmgexp[grep("m",mydf$propdmgexp)]<-6
     mydf$propdmgexp<-as.numeric(mydf$propdmgexp)
     mydf$propdmgexp[is.na(mydf$propdmgexp)]<-0
     return(mydf)
}

prep_cropdmgexp<-function(mydf) {
     mydf$cropdmgexp<-tolower(mydf$cropdmgexp)
     mydf$cropdmgexp[grep("k",mydf$cropdmgexp)]<-3
     mydf$cropdmgexp[grep("b",mydf$cropdmgexp)]<-9
     mydf$cropdmgexp[grep("m",mydf$cropdmgexp)]<-6
     mydf$cropdmgexp<-as.numeric(mydf$cropdmgexp)
     mydf$cropdmgexp[is.na(mydf$cropdmgexp)]<-0
     return(mydf)
}

mydf<-prep_cropdmgexp(mydf) %>% prep_propdmgexp  

# Add three new column variables: year, propdamage_M, cropdamage_M
# using cleaned up codes for damage
mydf$year<-(year(mydf$bgn_date))
mydf$propdamage_M<-with(mydf,propdmg*10^(propdmgexp-6))
mydf$cropdamage_M<-with(mydf,cropdmg*10^(cropdmgexp-6))
 

```


2. **Analyze and limit data by cutoff year and timezone:** we use the aggregate function to total the **propdamage_K**, **cropdamage_K**, **fatalities**, and **injuries** columns by year, and we can easily see by looking at the resulting dataframe **total_by_year** with the ```View()``` command that the last 19 years have the largest numbers in each risk, when manually sorted, and crop damage appears to have been untracked earlier.  The year 1992 was smaller than quantities in other years, but we thought the 20 years from 1992-2011 was a pragmatic time frame to analyze.  
We also used the aggregate function to group the totals along the time zones.  We added columns to the **total_by_tz** to see the percent contribution of each time zone to the totals and could see in the ordered dataframe that the top 3 time zones account for well over 75% of the impact columns.  For completeness, we included the MST timezone.
Thus, we filtered the dataset down to four time zones (EST, CST, PST, MST) and years 1992 and beyond.


``` {r filter_year_time, cache=TRUE, echo=TRUE, message=FALSE}
#---------------------------------------------------------------------------------------
#
# DATA PROCESSING
#
# 2. **Analyze and limit data by cutoff year and timezone:**
#
#---------------------------------------------------------------------------------------

premelt<-melt(data=mydf,id.vars=c("evtype","year","time_zone"), 
              measure.vars=c("fatalities","injuries","propdamage_M","cropdamage_M"))

# these were used in barplot code shown in the Appendix, not used in report
var<-c('health','health','economic','economic')
names(var)<-c('fatalities','injuries','propdamage_M','cropdamage_M')
premelt$impact<-var[premelt$variable]
premelt$damage<-rep("damage",length(premelt[,1]))

precast_year<-dcast(premelt,year~variable,fun=sum)
# sort and show variables by year; last 20 years are worth focusing on
head(precast_year[order(-precast_year$propdamage_M),],30)

precast_tz<-dcast(premelt,time_zone~variable,fun=sum)
# sort and show variables by time zone; obviously 4 stand out
head(precast_tz[order(-precast_tz$propdamage_M),],10)

main_zones<-c('PST','EST','MST','CST')
data_recent_fourtz<-filter(mydf,year>=1992 & time_zone%in%main_zones)

```

3. **Clean/recategorize weather event types:** There were 985 unique event types in the original **sdata** dataframe, which was loaded raw without any filtering. After filtering the data down to the last 20 years and to four main time zones, there were still 820 unique event types. Through an iterative process of adding text recognition phrases and running them in the **prep_evtype()** function and looking at the largest contributing event types in the **total_by_type** aggregated dataframe, we reduced the unique event types to 313 before calculating which were the most significant event types.  We chose the all events that accounted for more than 2% of the total, for each of the four categories of impact. We then did a union operation of all these factors to come up with 12 unique weather events, and then filtered the data to only include these rows. 

``` {r filter_event, cache=TRUE, echo=TRUE, message=FALSE}

#---------------------------------------------------------------------------------------
#
# DATA PROCESSING
#
# 3. **Clean/recategorize weather event types:**
# we analyze a bit more to find the top event types per risk, which is the hardest
# task given the messy recordkeeping
#---------------------------------------------------------------------------------------

length(unique(sdata$EVTYPE))
length(unique(data_recent_fourtz$evtype))

# Run this first and search for poorly categorized/ redundant weather events
total_by_type_dirty<-aggregate(cbind(propdamage_M,cropdamage_M,injuries,fatalities) ~ evtype, 
                         data_recent_fourtz, sum)
head(total_by_type_dirty[order(-total_by_type_dirty$propdamage_M),],20)


prep_evtype<-function(mydf) {
        # consolidate categories of weather     
        mydf$evtype[grep("+tornado+",mydf$evtype)]<-"tornado"
        mydf$evtype[grep("+heavy rain+",mydf$evtype)]<-"heavy rain"
        mydf$evtype[grep("+heat+",mydf$evtype)]<-"heat"
        mydf$evtype[grep("+flood+",mydf$evtype)]<-"flood" 
        mydf$evtype[grep("+rip current+",mydf$evtype)]<-"rip current"
        
        mydf$evtype[grep("+wind+",mydf$evtype)]<-"wind"
        mydf$evtype[grep("+cold+",mydf$evtype)]<-"cold/freeze"
        mydf$evtype[grep("+hurricane+",mydf$evtype)]<-"hurricane/tropical storm"
        mydf$evtype[grep("+typhoon+",mydf$evtype)]<-"hurricane/tropical storm"
        mydf$evtype[grep("+tropical storm+",mydf$evtype)]<-"hurricane/tropical storm"
        
        mydf$evtype[grep("+snow+",mydf$evtype)]<-"snow"
        mydf$evtype[grep("+blizzard+",mydf$evtype)]<-"snow"
        mydf$evtype[grep("+ice storm+",mydf$evtype)]<-"snow"
        mydf$evtype[grep("+winter storm+",mydf$evtype)]<-"snow"
        mydf$evtype[grep("+freeze+",mydf$evtype)]<-"cold/freeze"
        mydf$evtype[grep("+frost+",mydf$evtype)]<-"cold/freeze"
        
        mydf$evtype[grep("+fire+",mydf$evtype)]<-"fire"
        mydf$evtype[grep("+high surf+",mydf$evtype)]<-"storm surge/high surf"
        mydf$evtype[grep("+storm surge+",mydf$evtype)]<-"storm surge/high surf"
        mydf$evtype[grep("+hail+",mydf$evtype)]<-"hail"
        return(mydf)
} 

df_type_clean<-prep_evtype(data_recent_fourtz)

total_by_type<-aggregate(cbind(propdamage_M,cropdamage_M,injuries,fatalities) ~ evtype, 
                         df_type_clean, sum)
head(total_by_type[order(-total_by_type$propdamage_M),],30)
length(unique(df_type_clean$evtype))

# now down to 313 unique event types
# let's see how many of these account for most of results

# - - - - - - - - - - - - - - - 

total_by_type$pctprop<-with(total_by_type, round(propdamage_M/sum(propdamage_M),2))
total_by_type$pctcrop<-with(total_by_type, round(cropdamage_M/sum(cropdamage_M),2))
total_by_type$pctfat<-with(total_by_type, round(fatalities/sum(fatalities),2))
total_by_type$pctinj<-with(total_by_type, round(injuries/sum(injuries),2))

crop_events<-filter(total_by_type,total_by_type$pctcrop>0.02)$evtype
prop_events<-filter(total_by_type,total_by_type$pctprop>0.02)$evtype
fat_events<-filter(total_by_type,total_by_type$pctfat>0.02)$evtype
inj_events<-filter(total_by_type,total_by_type$pctinj>0.02)$evtype

econ_risks<-union(fat_events,inj_events)
health_risks<-union(crop_events,prop_events)
risks<-union(econ_risks,health_risks)
risks

# filtering data down to these 12 risks

final_data<-filter(df_type_clean,evtype%in%risks)
length(unique(final_data$evtype))


```

4. **Melt and recast data:**  

At this point, it appeared prudent to break our 20 year period into five four-year periods, so as to average out spikes in the data from less regular weather events. We added a column **period**.  annual spikes in the weather we still did not have a strictly tidy dataset, because the four important observation variables (fatalities, injuries, property damage, crop damage) were located in columns.  Thus, we melted the data so that only one variable appeared per line, added a column specifying whether the observation was health or economic,and defined another column as **damage** so that body counts and dollar impact could be plotted along the same x-axis.  

``` {r melt_cast,  cache=TRUE, echo=TRUE, message=FALSE,out.width = '\\maxwidth'}
#---------------------------------------------------------------------------------------
# 
# DATA PROCESSING
# 4. **Melt and recasting data:**  
#
#---------------------------------------------------------------------------------------

myperiod<-c(rep('1992-5',4),rep('1996-1999',4),rep('2000-3',4),rep('2004-7',4),rep('2008-11',4))
names(myperiod)<-c(1992:2011)

final_data$period<-myperiod[as.character(final_data$year)]


# first we want to melt the df into tidy date, with one observation per row
# and introduce the concept of health and economic cost
# health = of type fatalities or injuries and measured by body count
# economic = of type property or crop damage and measured by USD Millions
# so we will have a long skinny dataframe where we have a year, timezone, type of event, 
# and only one type of the four costs above measured

melted<-melt(data=final_data,id.vars=c("evtype","period","time_zone"), 
             measure.vars=c("fatalities","injuries","propdamage_M","cropdamage_M"))

var<-c('health','health','economic','economic')
names(var)<-c('fatalities','injuries','propdamage_M','cropdamage_M')

melted$impact<-var[melted$variable]
melted$damage<-rep("damage",length(melted[,1]))

casted<-dcast(melted,evtype+time_zone+period+variable+impact~damage,fun=sum)


```

5. **Time period analysis:** We used the **ggplot**/ **geom_bar()** plots with free y-axis scales to plot the damage in 4 four barplots (four impact variables) over the five time periods (see Figure 1).  We customized the color palette of fill colors so that each weather event was matched to a unique color that, when possible,  made intuitive sense (e.g., flood was blue, snow was white).  We set customized the y-labels to be more aesthetically pleasing.

``` {r code_time,  cache=TRUE, echo=TRUE, message=FALSE,out.width = '\\maxwidth'}
#---------------------------------------------------------------------------------------
# 
# DATA PROCESSING
# 4. Time period plot 
# 
# Now we set up 5 periods of 4 years each, covering the 20 years from 1992-2011
# the named list myperiod acts as a dictionary for plotting functions for mapping year to period
#---------------------------------------------------------------------------------------

prop_total = sum(subset(casted,variable=="propdamage_M")$damage)
crop_total = sum(subset(casted,variable=="cropdamage_M")$damage)
fat_total = sum(subset(casted,variable=="fatalities")$damage)
inj_total = sum(subset(casted,variable=="injuries")$damage)
print(paste("Property damage $M total 1992-2011 from 12 main events:", prop_total))
print(paste("Property damage $M total 1992-2011 from 12 main events:", crop_total))
print(paste('Property damage is ~',round(prop_total/crop_total,0),'x larger'))

print(paste("Total fatalities from 1992-2011 from 12 main events:", fat_total))
print(paste("Total injuries from 1992-2011 from 12 main events:", inj_total))
print(paste('Injury total is ~',round(inj_total/fat_total,0),'x larger'))
 
mycolors= c('black','brown','blue', 'pink','red','green', 'orange','cyan',
            'white','purple','gray','yellow')

damage.labs<-c("Property Damage $M","Crop Damage $M","Fatalities","Injuries")
names(damage.labs)<-c("propdamage_M","cropdamage_M","fatalities","injuries")

theme_update(text = element_text(size=8))
 # for the main title

time_period_plot<-ggplot(casted,aes(period,y=damage,fill=evtype,order=variable)) + 
        geom_bar(stat='identity') + facet_grid(impact+variable ~.,scales="free_y",
                                               labeller=labeller(variable=damage.labs))+
        theme(axis.text.x=element_text(size=8,colour='black',angle=0))+
        ggtitle('Figure 1. Impact of Weather Events Over Time')+
        labs(fill="Weather Event", scale_fill_manual(values = mycolors))
time_period_plot+ scale_fill_manual(values = mycolors)

 
```

6. **Geographic (time zone) analysis:** Because our analysis is directed to national and local government decisionmakers, it made sense to also include a geographic analysis of the data across the four impact variables.  We used the stacked barplots with similar coloration scheme in Figure 2. 

``` {r code_timezone,  cache=TRUE, echo=TRUE, message=FALSE,out.width = '\\maxwidth'}

#---------------------------------------------------------------------------------------
# 
# DATA PROCESSING
# 5. Geographic plot (by time zone)  
# 
#---------------------------------------------------------------------------------------

# TIME ZONE BY IMPACT
time_zone_plot<-ggplot(casted,aes(time_zone,y=damage,fill=evtype,order=variable))+
        geom_bar(stat='identity')+
        facet_grid(impact+variable ~.,scales="free_y",
                   labeller=labeller(variable=damage.labs))+
        theme(axis.text.x=element_text(size=8,colour='black',angle=0))+
        ggtitle('Figure 2. Impact of Weather Events by Geographic Time Zone')+
        labs(fill="Weather Event", scale_fill_manual(values = mycolors))
theme_update(text = element_text(size=8))

time_zone_plot+ scale_fill_manual(values = mycolors)

``` 

Because most damage in the US from weather occurs in the coastal and central time zones, the scale on Figures 1 and 2 obscure the damage that occurs in the Mountain Standard Time zone.  Thus, we provide Figure 3 to inform local authorities in that time zone. 

``` {r mst_plot,  cache=TRUE, echo=TRUE, message=FALSE,out.width = '\\maxwidth'}

theme_update(text = element_text(size=8))

mycolors= c('black','brown','blue', 'pink','red','green', 'orange',
            'white','purple','gray','yellow')

damage.labs<-c("Property Damage $M","Crop Damage $M","Fatalities","Injuries")
names(damage.labs)<-c("propdamage_M","cropdamage_M","fatalities","injuries")
theme_update(text = element_text(size=8))

mst_plot<-ggplot(subset(casted,time_zone=='MST'),aes(period,y=damage,fill=evtype,order=variable)) + 
        geom_bar(stat='identity') + facet_grid(impact+variable ~.,scales="free_y",
                                               labeller=labeller(variable=damage.labs))+
        theme(axis.text.x=element_text(size=8,colour='black',angle=0))+
        ggtitle('Figure 3. Impact of Weather Events Over Time in MST Time Zone')+
        labs(fill="Weather Event", scale_fill_manual(values = mycolors))

# to actually change to my customized colors
mst_plot+ scale_fill_manual(values = mycolors)

```

## Results

The events most harmful to population health and with greatest economic consequences are as follows:

### Heat, flood, tornados were top 3 causes of death and injury, with wind and lightning also consistently big contributors

These five weather types were also the main causes of injury; however, tornados were a bigger component of injuries than heat or flooding.
Event types that are most harmful to population health
 
Fatalities and injuries are the harmful effects to population health.  For the time period, event types, and time zones analyzed, there was a total of more than 9,800 fatalities and more than 64,000 injuries.  Unlike the concentration in a few event types for economic impact, the health impact was more evenly distributed among event types.  In particular, heat, flood, tornados, wind and lightning were consistently the top causes of death, although the deathtoll from tornados grew over time while that from heat declined.

These five weather types were also the main causes of injury; however, tornados were a bigger component of injuries than heat or flooding.

In the MST timezone, much of which is at higher altitudes, wind, snow and lightning are the main risks to population health.

### Flood and hurricanes/tropical storms accounted for most of property damage and second and third causes of crop damage,while drought was number 1 cause of crop damage  

Economic consequences are measured by property damage and crop damage.  As can be seen from the scales of Figures 1 and 2 and the calculated totals for 1992-2011 for the top 12 event types in the top four time zones, property damage in dollar terms was ~$374 billion, which is nearly 8x the cumulative crop damage of under $46 billion.  

Flood and hurricanes/tropical storms accounted most of the property damage, concentrated in the 2004-2007 timeframe and in the EST, CST, and PST timezones.  Drought was a major problem in the CST and EST time zones for crop damage.  In the MST timezone, much of which is at higher altitudes, hail is the main cause of property and crop damage. 
